Linting Linearity in Core/System FC

### 2023-01-19 report.tex
rules
tweaks

\begin{mathpar}
  \begin{array}{lcl}
    s,t & ::= & \star\\
        & \mid & t \lolli s\\
        & \mid & t \oplus s\\
        & \mid & t \tensor s\\
        & \mid & t \with s\\
        & \mid & \bang t\\
   \end{array}

\and

  \begin{array}{lcl}
      M,N & ::= & \star \mid \llet{\star = M}{N} \\
        & \mid & u \\
        & \mid & \lambda u. M \mid M~N\\
        & \mid & \textrm{inl}~M \mid \textrm{inr}~M \mid
        \ccase{M}{\textrm{inl}~u_1 \to N_1 \mid \textrm{inr}~u_2 \to N_2}\\
        & \mid & M \tensor N \mid \llet{u_1 \tensor u_2 = M}{N}\\
        & \mid & M \with N \mid \textrm{fst}~M \mid \textrm{snd}~M \\
        & \mid & \bang M \mid \llet{!u = M}{N} \\
   \end{array}
\end{mathpar}

\begin{mathpar}
    \infer*[right=($\tensor I$)]
    {\Delta \vdash M : A \and \Delta' \vdash N : B}
    {\Delta , \Delta' \vdash (M \tensor N) : A \tensor B}

    \and
\end{mathpar}

\begin{mathpar}
    \infer*[right=($\tensor E$)]
    {\Delta \vdash M : A \tensor B \and \Delta', u{:}A, v{:}B \vdash N : C}
    {\Delta , \Delta' \vdash \llet{u \tensor v}{N} : C}
\end{mathpar}

\begin{mathpar}
    \infer*[right=($\lolli I$)]
    {\Delta , u{:}A \vdash M : B}
    {\Delta \vdash \lambda u. M : A \lolli B}

    \and

    \infer*[right=($\lolli E$)]
    {\Delta \vdash M : A \lolli B \and \Delta' \vdash N : A}
    {\Delta, \Delta' \vdash M~N : B}
\end{mathpar}

\begin{mathpar}
    \infer*[right=($\with I$)]
    {\Delta \vdash M : A \and \Delta \vdash N : B}
    {\Delta \vdash M \with N : A \with B}

    \and

    \infer*[right=($\with E_L$)]
    {\Delta \vdash M : A \with B}
    {\Delta \vdash \textrm{fst}~M : A}

    \and

    \infer*[right=($\with E_R$)]
    {\Delta \vdash M : A \with B}
    {\Delta \vdash \textrm{snd}~M : B}
\end{mathpar}

\begin{mathpar}
    \infer*[right=($\oplus I_L$)]
    {\Delta \vdash M : A}
    {\Delta \vdash \textrm{inl}~M : A \oplus B}

    \and

    \infer*[right=($\oplus I_R$)]
    {\Delta \vdash M : B}
    {\Delta \vdash \textrm{inr}~M : A \oplus B}

    \and

    \infer*[right=($\oplus E$)]
    {\Delta \vdash M : A \oplus B \and \Delta', w_1{:}A \vdash N_1 : C \and
    \Delta', w_2{:}B \vdash N_2 : C}
    {\Delta, \Delta' \vdash \ccase{M}{\textrm{inl}~w_1 \to N_1 \mid \textrm{inr}~w_2 \to N_2} : C}
\end{mathpar}

\begin{mathpar}
    \infer*[right=($\star I$)]
    { }
    {\Delta \vdash \star : \star}

    \and

    \infer*[right=($\star E$)]
    {\Delta \vdash M : \star \and \Delta' \vdash N : B}
    {\Delta, \Delta' \vdash \llet{\star = M}{N} : B}

\end{mathpar}

\begin{mathpar}
    \infer*[right=($\bang I$)]
    {\Gamma ; \cdot \vdash M : A}
    {\Gamma ; \cdot \vdash !M : !A}

    \and

    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta \vdash M : \bang A \and \Gamma, u{:}A ; \Delta' \vdash N : C}
    {\Gamma ; \Delta, \Delta' \vdash \llet{!u = M}{N} : C}
\end{mathpar}

@@

Another defining feature of Haskell is its powerful type system. Features such
as type classes, arbitrary-rank types, functional dependencies, GADTs, type
families and (now) even impredicative polymorphism\cite{10.1145/3408971} have become
commonplace in Haskell code enforcing application level invariants and program
correctness through the types. For example, consider the definition of
\textbf{head} in the standard library, a function which takes the first element
of a list by pattern matching on the list constructors.

@@

If the list is empty, we terminate the program with an error, otherwise, we take
the first element. This function is unsafe -- our program might crash if we use
it on an invalid input. A common technique in Haskell is to use more expressive
types to assert properties about the values and get rid of the impossible cases
(e.g. a \textbf{NonEmpty} is a list that can not be empty). A well liked motto is
"make invalid states unrepresentable". In this light, we will introduce
Generalized Algebraic Data Types (GADTs) and create a list type indexed by size
for which we can write a completely safe \textbf{head} function by expressing
that the size of the list must be at least one, at the type level.

% TODO: But what are ADTs? Perhaps expand Haskell introduction section to
% include recursive ADTs
\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are a very popular feature with
which users can define data types as they would common recursive algebraic data
types, but that allows explicit type signatures for the data constructors where
the result type may differ in the type parameters across branches. Pattern
matching against GADTs can introduce local type refinements, that is, pattern
matching against certain data constructor will refine the type information known
for typechecking in individual case alternatives. We develop the length-indexed
lists example without discussing the type system and type inference details of
GADTs which we explore later in~\ref{related-work-gadts}.

@@

First, we define the natural numbers inductively: a natural number is either
zero (\textbf{Z}) or a successor (\textbf{S}) of another natural number (e.g.
the successor of the successor of zero is a natural number). The following
definition creates the \textbf{Nat} type, the \textbf{Z} and \textbf{S} data
constructors which construct values, \emph{and} the \textbf{Z} and \textbf{S}
\emph{type} constructors which construct types. The latter two exist because of
\emph{DataKinds} language feature which promotes term-level constructors to
type-level constructors.

@@

The length-indexed list is defined inductively as either an empty list \emph{of
size zero}, or the construction of a list by appending a new element to an
existing list \emph{of size $m$} whose final size is $m+1$ (\textbf{S m}). The
list \textbf{Cons~2~(Cons~1~Nil)} has type \textbf{Vec~(S~(S~Z))~Int} because
\textbf{Nil~::~Vec~Z~Int} and \textbf{Cons~1~Nil~::~Vec~(S~Z)~Int}. GADTs make
possible different data type constructors being parametrized over different type
parameters as we do with \textbf{Vec}'s size parameter being different depending
on the constructor that constructs the list.
    To define the safe \textbf{head} function, we must specify the type of the
input list taking into account that the size must not be zero. To that effect,
the function takes as input a \textbf{Vec (S n) a}, that is, a vector with size
(n+1) for all possible n's, which makes it impossible to pass a list of type
\textbf{Vec Z a} to \textbf{head}.

@@

Pattern matching on the \textbf{Nil} constructor would result in an innacessible
pattern warning because it is impossible to pass a list that would match
\textbf{Nil} -- it would entail the list has size zero; however, it's impossible
to call \textbf{head} on a list of size zero!

Linear types, which allow for a finer control over the usage of certain resources at the type
level, are another example of an extension to Haskell's type system, besides the aforementioned type-level
features, that makes it even more expressive.

>>>>

\begin{mathpar}
  \begin{array}{lcl}
    s,t & ::= & \star\\
        & \mid & t \lolli s\\
        & \mid & t \oplus s\\
        & \mid & t \tensor s\\
        & \mid & t \with s\\
        & \mid & \bang t\\
   \end{array}
\and
  \begin{array}{lcl}
      M,N & ::= & \star \mid \llet{\star = M}{N} \\
        & \mid & u \\
        & \mid & \lambda u. M \mid M~N\\
        & \mid & \textrm{inl}~M \mid \textrm{inr}~M \mid
        \ccase{M}{\textrm{inl}~u_1 \to N_1 \mid \textrm{inr}~u_2 \to N_2}\\
        & \mid & M \tensor N \mid \llet{u_1 \tensor u_2 = M}{N}\\
        & \mid & M \with N \mid \textrm{fst}~M \mid \textrm{snd}~M \\
        & \mid & \bang M \mid \llet{!u = M}{N} \\
   \end{array}
\end{mathpar}

\[
  \begin{array}{c}
    \infer*[right=($\tensor I$)]
    {\Delta \vdash M : A \and \Delta' \vdash N : B}
    {\Delta , \Delta' \vdash (M \tensor N) : A \tensor B}
\qquad
    \infer*[right=($\tensor E$)]
    {\Delta \vdash M : A \tensor B \and \Delta', u{:}A, v{:}B \vdash N : C}
    {\Delta , \Delta' \vdash \llet{u \tensor v}{N} : C}
\\[1em]
    \infer*[right=($\lolli I$)]
    {\Delta , u{:}A \vdash M : B}
    {\Delta \vdash \lambda u. M : A \lolli B}
\qquad
    \infer*[right=($\lolli E$)]
    {\Delta \vdash M : A \lolli B \and \Delta' \vdash N : A}
    {\Delta, \Delta' \vdash M~N : B}
\\[1em]
    \infer*[right=($\with I$)]
    {\Delta \vdash M : A \and \Delta \vdash N : B}
    {\Delta \vdash M \with N : A \with B}
    \qquad
    \infer*[right=($\with E_L$)]
    {\Delta \vdash M : A \with B}
    {\Delta \vdash \textrm{fst}~M : A}
\qquad
    \infer*[right=($\with E_R$)]
    {\Delta \vdash M : A \with B}
    {\Delta \vdash \textrm{snd}~M : B}
\\[1em]
    \infer*[right=($\oplus I_L$)]
    {\Delta \vdash M : A}
    {\Delta \vdash \textrm{inl}~M : A \oplus B}

\qquad
    \infer*[right=($\oplus I_R$)]
    {\Delta \vdash M : B}
    {\Delta \vdash \textrm{inr}~M : A \oplus B}
\\[1em]

    \infer*[right=($\oplus E$)]
    {\Delta \vdash M : A \oplus B \and \Delta', w_1{:}A \vdash N_1 : C \and
    \Delta', w_2{:}B \vdash N_2 : C}
    {\Delta, \Delta' \vdash \ccase{M}{\textrm{inl}~w_1 \to N_1 \mid \textrm{inr}~w_2 \to N_2} : C}
    \\[1em]
    \infer*[right=($\star I$)]
    { }
    {\Delta \vdash \star : \star}
\quad    \infer*[right=($\star E$)]
    {\Delta \vdash M : \star \and \Delta' \vdash N : B}
    {\Delta, \Delta' \vdash \llet{\star = M}{N} : B}
\\[1em]
    \infer*[right=($\bang I$)]
    {\Gamma ; \cdot \vdash M : A}
    {\Gamma ; \cdot \vdash !M : !A}
\qquad    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta \vdash M : \bang A \and \Gamma, u{:}A ; \Delta' \vdash N : C}
    {\Gamma ; \Delta, \Delta' \vdash \llet{!u = M}{N} : C}
  \end{array}
  \]

@@

Another defining feature of Haskell is its powerful type system. Features such
as type classes, arbitrary-rank types, functional dependencies, GADTs, type
families and (more recently) even impredicative polymorphism\cite{10.1145/3408971} have become
commonplace in Haskell code, enforcing application level invariants and program
correctness through the types. For example, consider the definition of
\textbf{head} in the standard library, a function which takes the first element
of a list by pattern matching on the list constructors.

@@

Note that if the list is empty, we terminate the program with an error. This
function is unsafe -- our program might crash if we use it on an invalid input.
A common technique in Haskell is to use more expressive types to assert
properties about the values and get rid of the unsafe cases (e.g. a
\textbf{NonEmpty} is a list that can not be empty). A well liked motto is "make
invalid states unrepresentable". In this light, we will introduce Generalized
Algebraic Data Types (GADTs) and create a list type indexed by size for which we
can write a completely safe \textbf{head} function. This is done by expressing that the size
of the list must be at least one at the type level.

% TODO: But what are ADTs? Perhaps expand Haskell introduction section to
% include recursive ADTs
\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are a very popular feature with
which users can define data types as they would common recursive algebraic data
types, with the added abilitiy to give explicit type signatures for the data
constructors where the result type may differ in the type parameters across
branches. Pattern matching against GADTs can introduce local type refinements,
that is, pattern matching against certain data constructors will refine the type
information used for typechecking individual case alternatives. We develop
the length-indexed lists example without discussing the type system and type
inference details of GADTs which we later explore in~\ref{related-work-gadts}.

@@

First, we define the natural numbers inductively: a natural number is either
zero (\textbf{Z}) or a successor (\textbf{S}) of another natural number (e.g.
the successor of the successor of zero is a natural number). The following
definition creates the \textbf{Nat} type, the \textbf{Z} and \textbf{S} data
constructors which construct values, \emph{and} the \textbf{Z} and \textbf{S}
\emph{type} constructors which construct types. The latter two exist because of
the \emph{DataKinds} language feature which promotes term-level constructors to
type-level constructors.

@@

The length-indexed list is defined inductively as either an empty list \emph{of
size zero}, or the construction of a list by appending a new element to an
existing list \emph{of size $m$} whose final size is $m+1$ (\textbf{S m}). The
list \textbf{Cons~'a'~(Cons~'b'~Nil)} has type \textbf{Vec~(S~(S~Z))~Char}
because \textbf{Nil~::~Vec~Z~Char} and \textbf{Cons~'a'~Nil~::~Vec~(S~Z)~Char}.
GADTs make possible different data constructors being parametrized over
different type parameters as we do with \textbf{Vec}'s size parameter being
different depending on the constructor that constructs the list.

To define the safe \textbf{head} function, we must specify the type of the input
list taking into account that the size must not be zero. To that effect, the
function takes as input a \textbf{Vec (S n) a}, that is, a vector with size
(n+1) for all possible n's, which makes it impossible to pass a list of type
\textbf{Vec Z a} to \textbf{head}.

@@

Pattern matching on the \textbf{Nil} constructor would result in an innacessible
pattern warning because it is impossible to pass a list that would match
\textbf{Nil} -- it would entail the argument list has size zero; however, it's
impossible to call \textbf{head} on a list of size zero!

Linear types, which allow for a finer control over the usage of certain resources at the type
level, are another example of an extension to Haskell's type system
%, besides the aforementioned type-level features
that makes it even more expressive.
